---
title: "Homework 2"
author: "Songyao Wang"
date: "2026-10-04"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(janitor)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE)
```

# Problem 1

First, I import the NYC Transit dataset, retain the requested variables, and
convert the `entry` variable from a character variable to a logical variable.

```{r clean-transit-data}
transit_data =
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("NA", "")
  ) |>
  clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

The cleaned transit dataset contains information about subway entrances and
exits, including station name and line, station coordinates, routes served,
entry and vending status, entrance type, and ADA compliance. I standardized the
variable names, retained the requested variables, and converted `entry` from
`YES`/`NO` values to a logical variable.

The resulting dataset has `r nrow(transit_data)` rows and
`r ncol(transit_data)` columns. The dataset is not tidy because the routes
served by each station are spread across `route1` through `route11` instead of
being stored in a single route variable.

## Transit questions

```{r transit-summaries}
number_stations =
  transit_data |>
  distinct(station_name, line) |>
  nrow()

number_ada_stations =
  transit_data |>
  filter(ada) |>
  distinct(station_name, line) |>
  nrow()

proportion_entry_without_vending =
  transit_data |>
  filter(vending == "NO") |>
  summarize(
    proportion = mean(entry, na.rm = TRUE)
  ) |>
  pull(proportion)
```

There are `r number_stations` distinct stations when stations are identified
using both station name and line. Of these, `r number_ada_stations` stations are
ADA compliant.

Among station entrances and exits without vending,
`r scales::percent(proportion_entry_without_vending, accuracy = 0.1)` allow
entrance.

I next reshape the route columns so that route number and route name are
separate variables.

```{r tidy-routes}
transit_routes =
  transit_data |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(route_number)
  )

number_a_stations =
  transit_routes |>
  filter(route_name == "A") |>
  distinct(station_name, line) |>
  nrow()

number_a_ada_stations =
  transit_routes |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(station_name, line) |>
  nrow()
```

There are `r number_a_stations` distinct stations that serve the A train.
Among these stations, `r number_a_ada_stations` are ADA compliant.

# Problem 2

I import the three requested Trash Wheel sheets. The specified Excel ranges
omit the first non-data row and the unused note columns. I clean the variable
names, remove rows without dumpster-specific information, standardize the
month and year variables, and create a variable identifying each Trash Wheel.

```{r clean-trash-wheel-data}
mr_trash_wheel =
  read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N710"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    wheel = "Mr. Trash Wheel",
    month = str_trim(month),
    year = as.integer(year),
    sports_balls = as.integer(round(sports_balls))
  )

professor_trash_wheel =
  read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M135"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    wheel = "Professor Trash Wheel",
    month = str_trim(month),
    year = as.integer(year)
  )

gwynnda =
  read_excel(
    "data/202509 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L352"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    wheel = "Gwynnda",
    month = str_trim(month),
    year = as.integer(year)
  )

trash_wheel_data =
  bind_rows(
    mr_trash_wheel,
    professor_trash_wheel,
    gwynnda
  ) |>
  select(wheel, everything())
```

```{r trash-wheel-summaries}
professor_total_weight =
  trash_wheel_data |>
  filter(wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  ) |>
  pull(total_weight)

gwynnda_june_2022_cigarettes =
  trash_wheel_data |>
  filter(
    wheel == "Gwynnda",
    month == "June",
    year == 2022
  ) |>
  summarize(
    total_cigarette_butts = sum(
      cigarette_butts,
      na.rm = TRUE
    )
  ) |>
  pull(total_cigarette_butts)
```

The combined dataset contains `r nrow(trash_wheel_data)` dumpster-specific
observations from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Important
variables include wheel name, dumpster number, collection date, trash weight
and volume, and counts of litter such as plastic bottles, cigarette butts, and
plastic bags.

For the available data, Professor Trash Wheel collected
`r round(professor_total_weight, 2)` tons of trash. Gwynnda collected
`r scales::comma(gwynnda_june_2022_cigarettes)` cigarette butts in June 2022.

# Problem 3

## Baseline demographic data

The baseline file begins with a descriptive row, so I skip that row during
import. I also treat `.` as a missing value, clean the variable names, convert
sex and APOE4 carrier status to factors with descriptive labels, and remove
participants who had MCI at baseline.

```{r clean-baseline-data}
mci_baseline =
  read_csv(
    "data/MCI_baseline.csv",
    skip = 1,
    na = "."
  ) |>
  clean_names() |>
  filter(
    is.na(age_at_onset) |
      age_at_onset > current_age
  ) |>
  mutate(
    sex = case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male"
    ) |>
      factor(
        levels = c("Female", "Male")
      ),
    apoe4 = case_match(
      apoe4,
      0 ~ "Non-carrier",
      1 ~ "Carrier"
    ) |>
      factor(
        levels = c("Non-carrier", "Carrier")
      )
  )
```

```{r baseline-summaries}
number_recruited =
  nrow(mci_baseline)

number_developed_mci =
  mci_baseline |>
  filter(!is.na(age_at_onset)) |>
  nrow()

average_baseline_age =
  mean(mci_baseline$current_age)

proportion_women_apoe4 =
  mci_baseline |>
  filter(sex == "Female") |>
  summarize(
    proportion = mean(apoe4 == "Carrier")
  ) |>
  pull(proportion)
```

After applying the baseline eligibility requirement,
`r number_recruited` participants remain in the study. Of these,
`r number_developed_mci` developed MCI during follow-up.

The average baseline age was `r round(average_baseline_age, 2)` years. Among
women, `r scales::percent(proportion_women_apoe4, accuracy = 0.1)` were APOE4
carriers.

## Longitudinal amyloid data

The amyloid file also begins with a descriptive row, so I skip that row during
import. I reshape the visit-specific columns into long format so that each row
represents one observed biomarker measurement for one participant.

```{r clean-amyloid-data}
mci_amyloid =
  read_csv(
    "data/mci_amyloid.csv",
    skip = 1
  ) |>
  clean_names() |>
  pivot_longer(
    baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_42_40"
  ) |>
  mutate(
    time_years = case_match(
      visit,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8
    )
  ) |>
  drop_na(amyloid_42_40) |>
  select(
    id = study_id,
    time_years,
    amyloid_42_40
  )
```

The cleaned amyloid dataset contains `r nrow(mci_amyloid)` observed biomarker
measurements from `r n_distinct(mci_amyloid$id)` participants. Each row includes
a participant ID, the number of years since baseline, and the measured amyloid
beta 42/40 ratio.

## Comparing and combining the datasets

```{r compare-and-join-mci-data}
baseline_only =
  anti_join(
    mci_baseline,
    mci_amyloid,
    by = "id"
  )

amyloid_only =
  anti_join(
    mci_amyloid,
    mci_baseline,
    by = "id"
  ) |>
  distinct(id)

mci_complete =
  inner_join(
    mci_baseline,
    mci_amyloid,
    by = "id"
  ) |>
  arrange(id, time_years)

write_csv(
  mci_complete,
  "data/mci_complete.csv"
)
```

There are `r nrow(baseline_only)` eligible participants appearing only in the
baseline dataset and `r nrow(amyloid_only)` participants appearing only in the
amyloid dataset.

The combined dataset retains participants found in both sources. It contains
`r nrow(mci_complete)` observed participant visits from
`r n_distinct(mci_complete$id)` participants. It includes baseline demographic
characteristics, MCI onset information, follow-up time, and the corresponding
amyloid beta 42/40 measurement.

The final dataset is exported as `data/mci_complete.csv`.

